A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-06 podc_paper/algorithm.tex
minor edits
Rewording to save some space

\Eric{Reworded following parag after Hossein's comments of Jan 4}
To propagate operations from a node $v$'s children to $v$, $P$ first observes
the blocks in both of $v$'s children that are not already in $v$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $v$'s blocks array using a \op{CAS} instruction.
Following \cite{}, we call this a 3-step sequence a
\op{Refresh} on $v$. %(see Figure \ref{fig::propagstep}).
A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on $v$.
However, since a successful \op{Refresh} propagates multiple pending operations 
from $v$'s children to $v$,
we can prove that if two \op{Refresh}es by $P$ on $v$ fail,
then $P$'s operation has been propagated to $v$ by some other process, so $P$ can continue 
onwards towards the~root.

@@

\begin{figure}[tb]
\input{tournamentTree.pdf_t}
\caption{An example tournament tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still propagating towards the root.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
\end{figure}

\begin{figure}
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as before.
Root blocks also have the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure}

@@

Performing a double refresh at each node along the path from the leaf to the root ensures 
a block containing the operation is appended to the root before the operation termintes.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
Thus, we linearize operations according to the block they belong to in the root's blocks array.
We can choose how to order operations in the same block, since they must be concurrent.

@@

Each \node\ of the tournament tree has an infinite array called \fld{blocks}.
To simplify the code, each \node\ has an empty \block\ $B_0$ in \fld{blocks}[0].
We define $E(B_0)$ and $D(B_0)$ to be the empty sequence.
Each \node\ also has a \fld{head} index that stores the position in the \fld{blocks} array to be used
for the next attempt to append a \block\ to this \node.

@@

    \If{\var{v} is a leaf node}
        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \cmt{required enqueue is in \var{v}.\fld{left}}
            \State \fld{dir} \assign\ \fld{left}
        \Else
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
        \EndIf \label{endChooseDir}
        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t. $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$\label{getChildIndex}
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} \label{getRecurse}
    \EndIf
\EndFunction{GetEnqueue}
\end{algorithmic}
\caption{\label{pseudocode2}Queue implementation, continued.}
%\here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?  Not much; I tried it.}
\end{figure}

@@

Once a \op{Dequeue} has been appended to a block of the process's leaf and propagated to the root,
the \opemph{IndexDequeue} routine is used to find the \op{Dequeue}'s location in the root.
More precisely, \opa{IndexDequeue}{v, b, i}
computes the block in the root and the rank
within that block  of the \var{i}th \op{Dequeue} of the block \var{B} stored in \var{v}.\fld{blocks}[\var{b}].
Let $R\sub{\var{v}}$ be the successful \nf{Refresh} on node \var{v}
whose CAS wrote a block \var{B} into \var{v}.\fld{blocks}[\var{b}].
Let $par$ be \var{v}.\fld{parent}. Without loss of generality, assume 
\var{v} is the left child of $par$. Let
$R_{par}$ be the first successful \op{Refresh} on $par$ whose \op{CreateBlock} reads a
value greater than \var{b} from $\var{par}.\fld{left.head}=\var{v}.\fld{head}$ at line \ref{createEndLeft} and therefore creates a block $B'$ that contains
\var{B} as a subblock. 
%Let $b'$ be the index of $B'$ in $par$.\fld{blocks}.

@@

To compute the response of the $i$th \nf{Dequeue} in the $b$th block
of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
If the queue is
not empty, Line \ref{computeE} computes the rank \var{e} of the
\nf{Enqueue} whose argument is the response to the
\nf{Dequeue}. 
A binary search is used on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} to find the index $b_e$ of the block that contains 
the \var{e}th enqueue.
The index $b_e$ is at most $b$ (since the enqueue is linearized before the dequeue).  To find the left end of the range for the binary search for $b_e$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
Then, \op{GetEnqueue} is used to trace down through the tree to find the required enqueue in a leaf.

>>>>

\Eric{Reworded following parag after Hossein's comments of Jan 4}
To propagate operations from a node $v$'s children to $v$, $P$ first observes
the blocks in both of $v$'s children that are not already in $v$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $v$'s blocks array using a \op{CAS} instruction.
Following \cite{DBLP:conf/fsttcs/JayantiP05}, we call this a 3-step sequence a
\op{Refresh} on $v$. %(see Figure \ref{fig::propagstep}).
A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on $v$.
However, since a successful \op{Refresh} propagates multiple pending operations 
from $v$'s children to $v$,
we can prove that if two \op{Refresh}es by $P$ on $v$ fail,
then $P$'s operation has been propagated to $v$ by some other process, so $P$ can continue 
onwards towards the~root.

@@

\begin{figure}[t]
\input{tournamentTree.pdf_t}
\caption{An example tournament tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still propagating towards the root.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
\end{figure}
\here{Make sure two figures appear on same page of final version}

\begin{figure}[t]
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{tournament}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as before.
Root blocks also have the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure}

@@

Performing a double refresh at each node along the path from the leaf to the root ensures 
a block containing the operation is appended to the root before the operation completes.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in the root's blocks array.
Thus, we linearize operations according to the block they belong to in the root's array.
We can choose how to order operations in the same block, since they must be concurrent.

@@

Each tournament tree \node\ has an infinite array called \fld{blocks}.
To simplify the code, each \node\ has an empty \block\ $B_0$ in \fld{blocks}[0].
We define $E(B_0)$ and $D(B_0)$ to be the empty sequence.
Each \node's \fld{head} index  stores the position in the \fld{blocks} array to be used
for the next attempt to append a \block.

@@

    \If{\var{v} is a leaf node}
        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \Comment{required enqueue is in \var{v}.\fld{left}}
            \State \fld{dir} \assign\ \fld{left}
        \Else
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
        \EndIf \label{endChooseDir}
        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t. $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$\label{getChildIndex}
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} \label{getRecurse}
    \EndIf
\EndFunction{GetEnqueue}
\end{algorithmic}
\caption{\label{pseudocode2}Queue implementation, continued.}
%\here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?  Not much; I tried it.}
\end{figure}

@@

Once a \op{Dequeue} is appended to a block of the process's leaf and propagated to the root,
the \opemph{IndexDequeue} routine finds the \op{Dequeue}'s location in the root.
More precisely, \opa{IndexDequeue}{v, b, i}
computes the block in the root and the rank
within that block  of the \var{i}th \op{Dequeue} of the block \var{B} stored in \var{v}.\fld{blocks}[\var{b}].
Let $R\sub{\var{v}}$ be the successful \nf{Refresh} on node \var{v}
whose CAS wrote a block \var{B} into \var{v}.\fld{blocks}[\var{b}].
Let $par$ be \var{v}.\fld{parent}. Without loss of generality, assume 
\var{v} is the left child of $par$. Let
$R_{par}$ be the first successful \op{Refresh} on $par$ whose \op{CreateBlock} reads a
value greater than \var{b} from $\var{par}.\fld{left.head}=\var{v}.\fld{head}$ at line \ref{createEndLeft} and therefore creates a block $B'$ that contains
\var{B} as a subblock. 
%Let $b'$ be the index of $B'$ in $par$.\fld{blocks}.

@@

To compute the response of the $i$th \nf{Dequeue} in the $b$th block
of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
If the queue is
not empty, Line \ref{computeE} computes the rank \var{e} of the
\op{Enqueue} whose argument is the \op{Dequeue}'s response. 
A binary search on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} finds the index $b_e$ of the block that contains 
the \var{e}th enqueue.
Since the enqueue is linearized before the dequeue, $b_e\leq b$.  To find the left end of the range for the binary search for $b_e$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
Then, \op{GetEnqueue} is used to trace down through the tree to find the required enqueue in a leaf.

### 2023-01-06 podc_paper/analysis.tex
Rewording to save some space

\begin{lemma}\label{enqDeqTime}
Each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
\Eric{In the thesis, Lemma 52 said enqueues take $O(\log^2 p)$ steps.  Was this a typo?}
and each non-null \op{Dequeue} takes
$O(\log p\log c + \log q_e+ \log q_d)$ steps,
where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and 
$q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.
\end{lemma}
\begin{proof}
An \op{Enqueue} or null \op{Dequeue} creates a block, appends it to the process's 
leaf and propagates it to the root.  The \op{Propagate} routine does $O(1)$ steps 
at each node along the path from the process's leaf to the root.
A null \op{Dequeue} additionally calls \op{IndexDequeue}, which also does $O(1)$ steps
at each node along this path. 
So, the total number of steps for either type of operation is $O(\log p)$.

A non-null \op{Dequeue} must, in addition, do a search at line \ref{FRb} and call \op{GetEnqueue}
at line \ref{findAnswer}.
\Eric{Modified:}
By \Cref{dSearchTime}, the search takes $O(\log(q_e+q_d+p))$ steps,
since the size of the queue can change by at most $p$ within one block of operations
(by \Cref{blockSize}).
\op{GetEnqueue} does a binary search within each node along a path from the root to a leaf.
The search in each node $v$ is within the subblocks of one block in $v$'s parent.
By \Cref{blockSize}, each search takes $O(\log c)$ time, for a total of $O(\log p\log c)$ steps.
\end{proof}

>>>>

\begin{lemma}\label{enqDeqTime}
Each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
\Eric{In the thesis, Lemma 52 said enqueues take $O(\log^2 p)$ steps.  Was this a typo?}
and each non-null \op{Dequeue} takes
$O(\log p\log c + \log q_e+ \log q_d)$ steps,
where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and 
$q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.
\end{lemma}
\begin{proof}
An \op{Enqueue} or null \op{Dequeue} creates a block, appends it to the process's 
leaf and propagates it to the root.  The \op{Propagate}  does $O(1)$ steps 
at each node on the path from the process's leaf to the root.
A null \op{Dequeue} additionally calls \op{IndexDequeue}, which also does $O(1)$ steps
at each node on this path. 
So, the total number of steps for either type of operation is $O(\log p)$.

A non-null \op{Dequeue} must also search at line \ref{FRb} and call \op{GetEnqueue}
at line \ref{findAnswer}.
\Eric{Modified:}
By \Cref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,
since the size of the queue can change by at most $p$ within one block (by \Cref{blockSize}).
\op{GetEnqueue} does a binary search within each node on a path from the root to a leaf.
Each node $v$'s search is within the subblocks of one block in $v$'s parent.
By \Cref{blockSize}, each search takes $O(\log c)$ steps, for a total of $O(\log p\log c)$ steps.
\end{proof}

### 2023-01-06 podc_paper/boundedspace.tex
Rewording to save some space

We replace the \fld{blocks} array in each node by a red-black tree (RBT)
that stores the blocks.
Each block has an additional \fld{index} field
that represents its position within the original \fld{blocks} array, and
blocks in a RBT are sorted by \fld{index}.
Accessing the block in entry $i$ in the \fld{blocks} array can then be 
replaced by searching the RBT for the block with index $i$.
The attempt to append a new block in entry $i$ of the \fld{blocks} array on line \ref{cas}
is replaced by an attempt to insert a new block with index $i$ into the RBT.

@@

To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
If a process performing a \op{Refresh} on the root finds that $root.\head$ is a multiple
of $p^2$, it does GC.
To determine which blocks can be thrown away, we keep an array $\var{last}[1..p]$ where 
each process writes the index of the last block in
\var{root} containing an enqueue whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending dequeue 
by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\var{root}.\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the root's RBT 
with indices smaller than $m$.
Fortunately, there is an efficient \op{Split} operation that can remove
all nodes with indices less than $m$ from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
Then, GC can work its way down the tournament tree, using \op{Split} operations to discard all subblocks
of the discarded root blocks.

@@

Because of GC, an operation may not find a block $B$ that it requires when it searches a RBT.  If the operation
is a dequeue, it must have been helped before $B$ was discarded, so it can simply read its response from
its own leaf.
\here{What about enqueues not finding required block?}

Each process takes $\tilde{O}(p)$ to run GC.\footnote{We use $\tilde{O}$ notation
to omit polylogarithmic factors; see Appendix \ref{reducing-details} for precise bounds.}
Since all processes may run GC,
a total of $\tilde{O}(p^2)$ steps is spent on a GC phase.
Between two of these GC phases, there are at least $p^2$ operations, so
GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.

>>>>

We replace the \fld{blocks} array in each node by a red-black tree (RBT)
that stores the blocks.
Each block has an additional \fld{index} field
that represents its position within the original \fld{blocks} array, and
blocks in a RBT are sorted by \fld{index}.
The attempt to append a new block in entry $i$ of the \fld{blocks} array on line \ref{cas}
is replaced by an attempt to insert a new block with index $i$ into the RBT.
Accessing the block in entry $i$ in the \fld{blocks} array can then be 
replaced by searching the RBT for the block with index $i$.
The binary searches for a block in line \ref{FRb} and \ref{getChild} can simply search the RBT
using the \fld{sum\sub{enq}} field, since the RBT is also sorted with respect to this field, by \Cref{lem::sum}.

@@

To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
If a process performing a \op{Refresh} on the root finds that $\var{root}.\head$ is a multiple
of $p^2$, it does GC.
To determine which blocks can be thrown away, we keep an array $\var{last}[1..p]$ where 
each process writes the index of the last block in
\var{root} containing an enqueue whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending dequeue 
by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\var{root}.\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the root's RBT 
with indices smaller than $m$.
Fortunately, there is an efficient \op{Split} operation that can remove
all nodes with indices less than $m$ from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
Then, GC can work its way down the tournament tree, using \op{Split} operations to discard all subblocks
of the discarded root blocks.

@@

An operation $op$'s search of a RBT may fail to find the required block $B$ that has been removed 
by another process's GC phase.  If $op$
is a dequeue, $op$ must have been helped before $B$ was discarded, so $op$ can simply read its response from
its own leaf.
\here{What about enqueues not finding required block?}

Each process takes $\tilde{O}(p)$ steps during a GC phase.\footnote{We use $\tilde{O}$ notation
to omit polylogarithmic factors; see Appendix \ref{reducing-details} for precise bounds.}
Since all processes may run GC,
a total of $\tilde{O}(p^2)$ steps is spent on a GC phase.
There are at least $p^2$ operations between two GC phases,  so
GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.

### 2023-01-06 podc_paper/conclusion.tex
Rewording to save some space

Our focus in constructing our queue was on optimizing step complexity for worst-case executions.
However, our approach does have a higher cost in the best case (for example, when an operation
runs in isolation).
It may be possible to make our queue adaptive by having an operation capture a starting node
in the tournament tree (as in \cite{??}) rather than having a statically assigned leaf as its starting point.
Another way to make our queue more practical might be to use it as the slow path in the
fast-path slow-path methodology of Kogan and Petrank \cite{10.1145/2370036.2145835} to
get a queue that has good performance in practice while also having good worst-case step complexity.

@@

We believe the approach used here to implement a lock-free queue 
could also be applied to obtain other lock-free
data structures with a polylogarithmic number of steps per operation.
For example, we could implement a kind of vector data structure that stores a sequence and
provides three operations: \opa{append}{e} to add an element \var{e} to the end of the sequence,
\opa{get}{i} to read the \var{i}th element in the sequence, and
\opa{index}{e} to compute the position of element \var{e} in the sequence.
\here{Can omit rest of this paragraph to save space}
An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
A \opa{get}{i} calls \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array, taking $O(\log n + \log^2p)$ when the vector has $n$ elements.  
An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and takes $O(\log p)$ steps.
\here{doublecheck all preceding claims for accuracy}.
As future work we would like to investigate whether a similar approach could be used for stacks or deques.

>>>>

Our focus in constructing our queue was on optimizing step complexity for worst-case executions.
However, our approach does have a higher cost in the best case (for example, when an operation
runs in isolation).
It may be possible to make our queue adaptive by having an operation capture a starting node
in the tournament tree (as in \cite{DBLP:conf/stoc/AfekDT95}) rather than having a statically assigned leaf as its starting point.
Another way to make our queue more practical might be to use it as the slow path in the
fast-path slow-path methodology of Kogan and Petrank \cite{10.1145/2370036.2145835} to
get a queue that has good performance in practice while also having good worst-case step complexity.

@@

We believe the approach used here to implement a lock-free queue 
could be applied to obtain other lock-free
data structures with a polylogarithmic number of steps per operation.
For example, we could implement a  vector data structure that stores a sequence and
provides three operations: \opa{append}{e} to add an element \var{e} to the end of the sequence,
\opa{get}{i} to read the \var{i}th element in the sequence, and
\opa{index}{e} to compute the position of element \var{e} in the sequence.
\here{Can omit rest of this paragraph to save space}
An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
A \opa{get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps.
\here{doublecheck all preceding claims for accuracy.}
We would like to investigate whether a similar approach could be used for stacks, deques or even priority queues.

### 2023-01-06 podc_paper/correctness.tex
Rewording to save some space

If $v$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
For line \ref{enqNew}, $B$ represents a single enqueue operation, so $|E(B)|=1$ and $|D(B)|=0$.
Moreover, $B.\fld{sum\sub{enq}}$ is set to $v.\fld{blocks}[i-1]+1$ and
$B.\fld{sum\sub{deq}}$ is set to $v.\fld{blocks}[i-1]$, so the claim follows from the hypothesis.
The proof for line \ref{deqNew}, where $B$ represents a single dequeue operation is similar.

@@

\begin{lemma} \label{lem::appendExactlyOnce}
When an \op{Append}($B$) terminates, $B$'s operation is contained in exactly one block in each node along the path from the process's leaf to the root.
\end{lemma}
\begin{proof}
The \op{Append} adds $B$ to the process's leaf and calls \op{Propagate}, which
performs a double \op{Refresh} on each node along the path from the leaf's parent to the root.
By \Cref{lem::doubleRefresh}, this ensures $B$'s operation is contained in each node along this path.
By \Cref{lem::noDuplicates}, it is contained in exactly one block in each node.
\end{proof}

>>>>

If $v$ is a leaf, $B$ was created at line \ref{enqNew} or \ref{deqNew}.
For line \ref{enqNew}, $B$ represents a single enqueue, so $|E(B)|=1$ and $|D(B)|=0$.
Since $B.\fld{sum\sub{enq}}$ is set to $v.\fld{blocks}[i-1]+1$ and
$B.\fld{sum\sub{deq}}$ is set to $v.\fld{blocks}[i-1]$, the claim follows from the hypothesis.
The proof for line \ref{deqNew}, where $B$ has a single dequeue, is similar.

@@

\begin{lemma} \label{lem::appendExactlyOnce}
When an \op{Append}($B$) terminates, $B$'s operation is contained in exactly one block in each node along the path from the process's leaf to the root.
\end{lemma}
\begin{proof}
\op{Append} adds $B$ to the process's leaf and calls \op{Propagate}, which
does a double \op{Refresh}~on each internal node on the path $P$ from the leaf to the root.
By \Cref{lem::doubleRefresh}, this ensures a block in each node on $P$ contains $B$'s operation.
There is at most one such block in each node, by \Cref{lem::noDuplicates}.
\end{proof}
